Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2021-05-19 paper_io_sim.tex
Pass on the whole paper with mostly minor stuff Swapped the order of two paragraphs in Section II Added \todo for comments/questions/thoughts

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{graphicx}
\usepackage{paralist}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage[dvipsnames]{xcolor}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{multirow}

@@

    The emergence of Big Data in recent years has resulted in a growing
    need for efficient data processing solutions. While infrastructures
    with sufficient compute power are available,
    the I/O bottleneck remains. The Linux page cache is an efficient
    approach to reduce I/O overheads, but few
    experimental studies of its interactions with Big Data applications exist,
    partly due to limitations of
    real-world experiments. Simulation is a popular approach to address
    these issues, however, existing simulation frameworks do not simulate
    page caching fully, or even at all.  As a result, simulation-based
    performance studies of data-intensive applications lead to inaccurate
    results.

    In this paper, we propose an I/O simulation model that includes
    the key features of the Linux page cache. We have implemented this model
    as part of the \wrench workflow simulation framework, which itself
    builds on the popular \simgrid distributed systems simulation
    framework. Our model and its implementation enable the simulation
    of both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems. We evaluate the accuracy of our model in different
    conditions, including sequential and concurrent applications, as
    well as local and remote I/Os. We find that our page cache model
    reduces the simulation error by up to an order of magnitude when
    compared to state-of-the-art, cacheless simulations.

@@

        The Linux page cache plays an important role in
        reducing filesystem data transfer times. With the page cache, previously
        read data can be re-read directly from memory, and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The performance improvements
        depend on many factors including the total amount of memory,
        the amount of data being written (i.e., dirty data), and the amount of memory available for
        written data. All these factors are important when determining the impact of I/O on
        application performance, particularly in data-intensive applications.

        % Swapped the order the two paragraphs and merged them into one
        The number of data-intensive applications has been steadily rising as a result of
        open-data and data sharing initiatives. Due to the sheer size of the data being
        processed, these applications must be executed on large-scale infrastructures
    such as High Performance Computing (HPC) clusters or the cloud.  It
    is thus crucial to quantify the performance of these applications
    on these platforms. The goals include determining which type of hardware/software
    stacks are best suited to different application classes, as well as
    understanding the limitations of current algorithms, designs and
    technologies. Unfortunately, performance
        studies relying on real-world experiments on compute platforms
        face several difficulties (high operational costs, labor-intensive experimental setups,
        shared platforms with dynamic loads that hinder reproducibility of results) and shortcomings
        (experiments are limited to the available platform/software configurations, which precludes
        the exploration of hypothetical scenarios). 
    Simulations address these concerns by providing models
    and abstractions for the performance of computer hardware, such as
    CPU, network and storage. As a result, simulations provide a
    cost-effective, fast, easy and reproducible way to evaluate
    application performance on arbitrary platform configurations. It thus comes
        as no surprise that a large number of simulation frameworks have been developed and used 
        for research and development~\cite{ optorsim, gridsim, groudsim, cloudsim, nunez2012simcan,nunez2012icancloud, mdcsim, dissect_cf, cloudnetsimplusplus, fognetsimplusplus, casanova2014simgrid, ROSS, casanova2020fgcs}.

        Page caching is an ubiquitous technique for mitigating the I/O bottleneck.
        As such, it is necessary to model it
        when simulating data-intensive applications.
        While existing simulation frameworks of parallel and distributed computing
        systems  capture many relevant features of hardware/software stacks, 
        they lack the ability to simulate page cache with enough details to capture key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Some simulators, such as the one in~\cite{xu2018saving}, do capture such features,
        but are domain-specific.

@@

    In this work, we present \wrench-cache, a page cache simulation model
    implemented in \wrench~\cite{casanova2020fgcs}, a workflow simulation
    framework based on the popular \simgrid distributed simulation
    toolkit~\cite{casanova2014simgrid}. Our contributions are:
    \begin{compactitem}
        \item A page cache simulation model that supports 
    both single-threaded and multithreaded applications, and both
    writeback and writethrough caches for local or network-based
    filesystems;
        \item An implementation of this model in \wrench; 
        \item An evaluation of the accuracy and scalability of our model, and of its implementation,
              for multiple applications, execution scenarios, and page cache configurations. 
    \end{compactitem}

    \section{Related Work}
    \label{relatedwork}

@@

        Cache eviction and flushing strategies are integral to proper page cache functioning.
        Whenever space in memory becomes limited, either as a result of application memory
        or page cache use, page cache data may be evicted. Only data that
        has been persisted to storage (clean pages) can be flagged for eviction and removed from
        memory. Written data that has not yet been persisted to disk (dirty data) must first
        be copied (flushed) to storage prior to eviction. When sufficient memory is
        being occupied, the flushing process is synchronous. However, even when
        there is sufficient available memory, written data will be flushed to disk
        at a predefined interval through a process known as \textit{periodical flushing}.
        Periodical flushing only flushes expired dirty pages, which remain dirty in
        page cache longer than an expiration time configured in the kernel.
        Different cache eviction algorithms have also been proposed
        \cite{owda2014comparison}.

@@

        In this  work, we use the \simgrid and \wrench simulation
        frameworks.  The years of research and development invested in
        the popular \simgrid simulation framework~\cite{casanova2014simgrid}, have
        culminated in a set of state-of-the-art macroscopic simulation
        models that yield high accuracy, as demonstrated by
        (in)validation studies and comparisons to competing
        frameworks~\cite{smpi_validity, velhoTOMACS2013, simutool_09,
        nstools_07, lebre2015, pouilloux:hal-01197274,
        smpi_tpds2017,  7885814, 8048921, 7384330}.  But one
        significant drawback of \simgrid is that its simulation
        abstractions are low-level, meaning that implementing a
        simulator of complex systems can be
        labor-intensive~\cite{kecskemeti_2014}. To remedy this problem,
        the \wrench simulation framework~\cite{casanova2020fgcs}
        builds on top of \simgrid to provide higher-level simulation
        abstractions, so that simulators of complex applications and
        systems can be implemented with a few hundred lines.

        Although the Linux page cache has a large impact on I/O
        performance, and thus on the execution of data-intensive
        applications, its simulation is rarely considered in the above
        frameworks.  Most frameworks merely simulate I/O operations
        based on storage bandwidths and capacities.  The SIMCAN
        framework does models page caching by storing data accessed on
        disk in a block cache~\cite{nunez2012simcan}.  Page cache is
        also modeled in iCanCloud through a component that manages
        memory accesses and cached data~\cite{nunez2012icancloud}.
        However, the scalability of the iCanCloud simulator is limited
        as it uses microscopic models.  Besides, none
        of these simulators provide any writeback cache simulator nor
        cache eviction policies through LRU lists.  Although cache
        replacement policies are applied in~\cite{xu2018saving} to
        simulate in-memory caching, this simulator is specific to
        energy consumption of multi-tier heterogeneous networks.

        In this study, we implement a page cache simulation model in the
        \wrench framework. We targeted \wrench because it is a recent,
        actively developed framework that provides convenient simulation
        abstractions, because it is extensible, and because it reuses
        \simgrid's scalable and accurate models.

    \section{Methods}
    \label{method}
    We separate our simulation model in two components, the I/O
    Controller and the Memory Manager, which together simulate
    file reads and writes (Figure~\ref{fig:interaction}).
    To read or write a file chunk, a simulated application sends a
    request to the I/O Controller. The I/O Controller interacts as needed with
    the Memory Manager to free memory through flushing or eviction,
    and to read or write cached data. The Memory Manager
    implements these operations, simulates periodical flushing
    and eviction, and reads or writes to disk when necessary.
    In case the writethrough strategy is used, the I/O Controller directly writes to disk, 
    cache is flushed if needed and written data is added to page cache.

@@

    Our simulation model supports chunk-by-chunk file accesses
    with a user-defined chunk size. However, for simplicity, we assume that file pages are
    accessed in a round-robin fashion rather than fully randomly.
    Therefore, when a file is read, cached data is read only after all uncached data was read, and data from the inactive list is read
    before data from the active list
    (data reads occur from left to right in Figure~\ref{fig:read_order}).
    When a chunk of \emph{uncached} data is read, a new clean block is created
    and appended to the inactive list.
    When a chunk of \emph{cached} data is read, one or more existing data blocks in the LRU lists are accessed.
    If these blocks are clean, we merge them together, update the access time and size of the resulting block,
    and append it to the active list.
    If the blocks are dirty, we move them independently to the active list, to preserve their entry time.
    Because the chunk and block sizes may be different, there are situations
    where a block is not entirely read.
    In this case, the block is split in two smaller blocks and one of them is re-accessed.
    % From Tristan: not sure where to put this nor if it's necessary:
    % , in which chunks are read/written
    % until file is entirely read/written.
    \begin{figure}
           \centering
           \includegraphics[width=\columnwidth]{figures/read_order.pdf}
           \caption{File data read order. Data is read from left to right: uncached data
           is read first, followed by data from the inactive list, and finally data from the active list. }
           \label{fig:read_order}
    \end{figure}

@@

    The main simulated thread in the Memory Manager can flush or evict data from the
    memory cache. The data flushing simulation
    function takes the amount of data to flush as parameter. While
    this amount is not reached and dirty
    blocks remain in cache, this function traverses the sorted
    inactive list, then the sorted active list, and writes the
    least recently used dirty block to disk, having set its dirty
    flag to 0. In case the amount of data to flush requires that a
    block be partially flushed, the block is split in two blocks,
    one that is flushed and one that remains dirty. The time needed
    to flush data to disk is simulated by the storage model.

    The cache eviction simulation also runs in
    the main thread. It frees up the page cache by traversing and deleting
    least recently used clean data blocks in the inactive list.
    The amount of data to evict is passed as a parameter and data blocks are deleted
    from the inactive list until the evicted data reaches the required amount,
    or until there is no clean block left in the list.
    If the last evicted block does not have to be entirely evicted, the block is split in two blocks,
    and only one of them is evicted.
    The overhead of the cache eviction algorithm is not part of the simulated time
    since cache eviction time is negligible in real systems. % \tristan{a reference would be welcome}.

@@

        \end{algorithmic}
    \end{algorithm}
    Algorithm~\ref{alg:write} describes our simulation of chunk writes in
    the I/O Controller.
    % Data is written to cache until the amount of dirty data in the cache
    % exceeds the dirty ratio times the amount of available memory. Data is then written to disk.
    Our algorithm initially checks the  amount of dirty data that
    can be written given the dirty ratio (line 5).
    If this amount is greater than 0, the Memory Manager is requested to evict
    data from cache if necessary (line 7).
    After eviction, the amount of data that can be written to
    page cache is calculated (line 8), and a cache write is simulated (line 9).
    If the dirty threshold is reached and there is still data to write,
    the remaining data is written to cache in a loop
    where we repeatedly flush and evict from the cache (line 12-18).

    The above model describes page cache in writeback
    mode. Our model also includes a write function in writethrough mode,
    which simply simulates a disk write with the amount of data passed in,
    then evicts cache if needed and adds the written data to the cache.

@@

            We first created a standalone prototype
            simulator to evaluate the accuracy and correctness of our
            model in a simple scenario before integrating it in the more complex
            \wrench framework.
            The prototype uses the following basic storage model for
            both memory and disk:
            \begin{align*}
                & t_{r} = D / b_r \\
                & t_{w} = D / b_w\
            \end{align*}

@@

        \subsection{Experiments}

        Our experiments compared real executions with our Python prototype,
        with the original \wrench simulator, and with our \wrench-cache
        extension. Executions included single-threaded and multi-threaded
        applications, accessing data on local and network file systems. We
        used two applications: a synthetic one, created to evaluate the
        simulation model, and a real one, representative of neuroimaging
        data processing.

        Experiments were run on a dedicated cluster at
        Concordia University, with one login node, 9 compute nodes, and 4
        storage nodes connected with a 25 Gbps network. Each
        compute node had 2 $\times$ 16-core Intel(R) Xeon(R) Gold 6130 CPU
        @ 2.10GHz, 250~GiB of RAM, 6 $\times$ SSDs of 450~GiB each with the XFS
        file system, 378~GiB of tmpfs, 126~GiB of devtmpfs file system,
        CentOS~8.1 and NFS version 4. We used the \texttt{atop}
        and \texttt{collectl} tools to monitor and collect memory status
        and disk throughput. We cleared the page
        cache before each application run to ensure comparable
        conditions.

        \begin{table}[b]
            \centering
            \begin{tabularx}{0.8\columnwidth}{c>{\centering\arraybackslash}X}
            \toprule
                Input size (GB)  & CPU time (s)\\
            \midrule
                3      & 4.4 \\
                20  & 28 \\
                50  & 75 \\
                75  & 110 \\
                100  & 155 \\
            \bottomrule
            \end{tabularx}
            \caption{Synthetic application parameters}
            \label{table:cputime}
            \end{table}
        The synthetic application, implemented in C, consisted of three single-core,
        sequential tasks where each task read the file produced by the
        previous task, incremented every byte of this file to emulate real
        processing, and wrote the resulting data to disk. Files were
        numbered by ascending access times (File 1 was the file read by Task 1, etc).
         The anonymous memory used by the application
        was released after each task, which we also simulated in the Python
        prototype and in WRENCH-cache. As our focus was on I/O rather than compute, we measured
        application task CPU times on a cluster node
        (Table~\ref{table:cputime}), and used these durations in our
        simulations. For the Python prototype, we injected CPU times
        directly in the simulation. For \wrench and \wrench-cache, we
        determined the corresponding number of flops on a 1~Gflops CPU
        and used these values in the simulation. The simulated
        platform and application are available at
        commit \href{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}{ec6b43561b}.

        We used the synthetic application in three experiments. In the
        first one (\textit{Exp~1}), we ran \emph{a single instance} of
        the application on a single cluster node, with different input file
        sizes (20~GB, 50~GB, 75~GB, 100~GB), and with all I/Os directed to
        the same local disk.
        In the second experiment (\textit{Exp~2}), we ran
        \emph{concurrent} instances of the application on a single node,
        all application instances operating on different files of size 3~GB
        stored in the same local disk. We varied the number of concurrent
        application instances from 1 to 32 since cluster nodes had 32 CPU
        cores.
        In the third experiment (\textit{Exp~3}), we used the same
        configuration as the previous one, albeit reading and writing
        on a 50-GiB NFS-mounted partition of a 450-GiB remote disk of
        another compute node. As is commonly configured in HPC
        environments to avoid data loss, there was no client write cache
        and the server cache was configured as writethrough instead of
        writeback. NFS client and server read caches were enabled. 
        Therefore, all the writes happened at disk bandwidth, but
        reads could benefit from cache hits.

        The real application was a workflow of the Nighres
        toolbox~\cite{huntenburg2018nighres}, implementing cortical
        reconstruction from brain images in four steps: skull stripping,
        tissue classification, region extraction, and cortical
        reconstruction. Each step read files produced by the previous step,
        and wrote files that were or were not read by the subsequent step.
        More information on this application is available in the Nighres
        documentation at
        \url{https://nighres.readthedocs.io/en/latest/auto_examples/example_02_cortical_depth_estimation.html}.
        The application is implemented as a Python script that calls Java
        image-processing routines. We used Python 3.6, Java 8, and Nighres
        1.3.0. We patched the application to remove lazy data loading and
        data compression, which made CPU time difficult to separate from
        I/O time, and to capture task CPU times to inject them in the
        simulation. The patched code is available at
        \url{https://github.com/dohoangdzung/nighres}.
        \begin{table}[t]
            \centering
            \begin{tabular}{lccc}
            \toprule
                \multicolumn{1}{c}{Workflow step}& Input size       & Output size      & CPU time\\
                                       & (MB)             & (MB)             & (s)\\
            \midrule
               Skull stripping         &  295             & 393               & 137 \\
               Tissue classification   &  197              & 1376              & 614 \\
               Region extraction       &  1376             & 885              & 76 \\
               Cortical reconstruction &  393              & 786              & 272\\
            \bottomrule
            \end{tabular} 
            \caption{Nighres application parameters}
            \label{table:nighres_stats}
            \end{table}
            \begin{table}[b]
                \centering
                \begin{tabularx}{\columnwidth}{ll
                >{\centering\arraybackslash}X
                >{\centering\arraybackslash}X
                >{\centering\arraybackslash}X}
                \toprule
                    \multicolumn{2}{c}{Bandwidths}  & Cluster (real) & Python prototype & \wrench simulator\\
                \midrule
                \multirow{2}{*}{Memory}      & read  & 6860 & 4812 & 4812\\
                                             & write & 2764 & 4812 & 4812\\
                \multirow{2}{*}{Local disk}  & read  & 510  & 465  & 465\\
                                             & write & 420  & 465  & 465\\
                \multirow{2}{*}{Remote disk} & read  & 515  & -    & 445\\
                                             & write & 375  & -    & 445\\
                \multicolumn{2}{l}{Network}  & 3000  & -    & 3000\\
                \bottomrule
                \end{tabularx}
                \caption{Bandwidth benchmarks (MBps) and simulator configurations.
                The bandwidths used in the simulations were the average of the measured read and write bandwidths.
                Network accesses were not simulated in the Python prototype.}
                \label{table:benchmark}
                \end{table}

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{graphicx}
\usepackage{paralist}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage[dvipsnames]{xcolor}
\usepackage{todonotes}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{multirow}

@@

    The emergence of Big Data in recent years has resulted in a growing
    need for efficient data processing solutions. While infrastructures
    with sufficient compute power are available,
    the I/O bottleneck remains. The Linux page cache is an efficient
    approach to reduce I/O overheads, but few
    experimental studies of its interactions with Big Data applications exist,
    partly due to limitations of
    real-world experiments. Simulation is a popular approach to address
    these issues, however, existing simulation frameworks do not simulate
    page caching fully, or even at all.  As a result, simulation-based
    performance studies of data-intensive applications can lead to misleading
    results.

    In this paper, we propose an I/O simulation model that captures
    the key features of the Linux page cache. We have implemented this model
    as part of the \wrench workflow simulation framework, which itself
    builds on the popular \simgrid distributed systems simulation
    framework. Our model and its implementation enable the simulation
    of both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems. We evaluate the accuracy of our model in different
    conditions, including sequential and concurrent applications, as
    well as local and remote I/Os. We find that our page cache model
    reduces the simulation error by up to an order of magnitude when
    compared to state-of-the-art, cacheless simulations.

@@

        The Linux page cache plays an important role in
        reducing filesystem data transfer times. With the page cache, previously
        read data can be re-read directly from memory, and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The performance improvements
        depend on many factors including the total amount of memory,
        the amount of data being written (i.e., dirty data), and the amount of memory available for
        written data. All these factors are important when determining the impact of I/O on
        application performance, particularly for data-intensive applications.

        % Swapped the order the two paragraphs and merged them into one
        The number of data-intensive applications has been steadily rising as a result of
        open-data and data sharing initiatives. Due to the sheer size of the data being
        processed, these applications must be executed on large-scale infrastructures
    such as High Performance Computing (HPC) clusters or the cloud.  It
    is thus crucial to quantify the performance of these applications
    on these platforms. The goals include determining which type of hardware/software
    stacks are best suited to different application classes, as well as
    understanding the limitations of current algorithms, designs and
    technologies. Unfortunately, performance
        studies that rely on real-world experiments on compute platforms
        face several difficulties (high operational costs, labor-intensive experimental setups,
        shared platforms with dynamic loads that hinder reproducibility of results) and have several shortcomings
        (experiments are limited to the available platform/software configurations, which precludes
        the exploration of hypothetical scenarios). 
    Simulation addresses these concerns by providing models
    and abstractions for the performance of computer hardware, such as
    CPU, network and storage. As a result, simulations provide a
    cost-effective, fast, convenient and reproducible way to evaluate
    application performance on arbitrary platform configurations. It thus comes
        as no surprise that a large number of simulation frameworks have been developed and used 
        for research and development in the field of parallel and distributed computing~\cite{ optorsim, gridsim, groudsim, cloudsim, nunez2012simcan,nunez2012icancloud, mdcsim, dissect_cf, cloudnetsimplusplus, fognetsimplusplus, casanova2014simgrid, ROSS, casanova2020fgcs}.

        Page caching is an ubiquitous technique for mitigating the I/O bottleneck.
        As such, it is necessary to model it
        when simulating data-intensive applications.
        While existing simulation frameworks of parallel and distributed computing
        systems  capture many relevant features of hardware/software stacks, 
        most lack the ability to simulate page cache with enough details to capture key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Some simulators, such as the one in~\cite{xu2018saving}, do capture such features
        but are domain-specific.

@@

    In this work, we present \wrench-cache, a page cache simulation model
    implemented in \wrench~\cite{casanova2020fgcs}, a workflow simulation
    framework that builds on the \simgrid distributed systems simulation
    framework~\cite{casanova2014simgrid}. Our contributions are:
    \begin{compactitem}
        \item A page cache simulation model that supports 
    both single-threaded and multithreaded applications, and both
    writeback and writethrough caches for local or network-based
    filesystems;
        \item An implementation of this model in \wrench; 
        \item An evaluation of the accuracy and scalability of our model, and of its implementation,
              for multiple applications, execution scenarios, and page cache configurations. 
    \end{compactitem}

    \section{Background and Related Work}
    \label{relatedwork}

@@

        Cache eviction and flushing strategies are integral to proper page cache functioning.
        Whenever space in memory becomes limited, either as a result of application memory
        or page cache use, page cache data may be evicted. Only data that
        has been persisted to storage (clean pages) can be flagged for eviction and removed from
        memory. Written data that has not yet been persisted to disk (dirty data) must first
        be copied (flushed) to storage prior to eviction. When sufficient memory is
        being occupied, the flushing process is synchronous. However, even when
        there is sufficient available memory, written data will be flushed to disk
        at a predefined interval through a process known as \textit{periodical flushing}.
        Periodical flushing only flushes ``expired" dirty pages, that is, pages that
        have remained dirty in the
        page cache longer than an expiration time configured in the kernel.
        Different cache eviction algorithms have also been proposed
        \cite{owda2014comparison}.

@@

        Although the Linux page cache has a large impact on I/O
        performance, and thus on the execution of data-intensive
        applications, its simulation is rarely considered in the above
        frameworks.  Most frameworks merely simulate I/O operations
        based on storage bandwidths and capacities.  The SIMCAN
        framework does models page caching by storing data accessed on
        disk in a block cache~\cite{nunez2012simcan}.  Page cache is
        also modeled in iCanCloud through a component that manages
        memory accesses and cached data~\cite{nunez2012icancloud}.
        However, the scalability of these simulators is limited
        as they rely on microscopic simulation models.  Besides, none
        of these simulators provide any writeback cache simulator nor
        cache eviction policies through LRU lists.  Although cache
        replacement policies are applied in~\cite{xu2018saving} to
        simulate in-memory caching, this simulator is specific to
        energy consumption of multi-tier heterogeneous networks.

        In this  work, we use the \simgrid and \wrench simulation
        frameworks.  The years of research and development invested in
        \simgrid~\cite{casanova2014simgrid}, have
        culminated in a set of state-of-the-art macroscopic simulation
        models that yield high accuracy, as demonstrated by
        (in)validation studies and comparisons to competing
        frameworks~\cite{smpi_validity, velhoTOMACS2013, simutool_09,
        nstools_07, lebre2015, pouilloux:hal-01197274,
        smpi_tpds2017,  7885814, 8048921, 7384330}.  One
        significant drawback of \simgrid is that its simulation
        abstractions are low-level, meaning that implementing a
        simulator of complex systems can be
        labor-intensive~\cite{kecskemeti_2014}. To remedy this problem,
        the \wrench simulation framework~\cite{casanova2020fgcs}
        builds on top of \simgrid to provide higher-level simulation
        abstractions, so that simulators of complex applications and
        systems can be implemented with a few hundred lines of code. 
        Neither \wrench nor \simgrid simulate page cache simulation.

	In this study, we implement a page cache simulation model in the
	\wrench framework. We target \wrench because it is a recent,
	actively developed framework that provides convenient and
	high-level simulation abstractions, because it is straightforward
	to extend, and because it builds on \simgrid's scalable and
	validated simulation models.

    \section{Methods}
    \label{method}
    \todo[inline]{HC: Personally, I would actually make this subsection a section, and rename the previous section from "Methods" to "Page Cache Simulation Model". I've never loved the "Methods" title for section, although I know it is standard practice for many authors. }

    We separate our simulation model in two components, the I/O
    Controller and the Memory Manager, which together simulate
    file reads and writes (Figure~\ref{fig:interaction}).
    To read or write a file chunk, a simulated application sends a
    request to the I/O Controller. The I/O Controller interacts as needed with
    the Memory Manager to free memory through flushing or eviction,
    and to read or write cached data. The Memory Manager
    implements these operations, simulates periodical flushing
    and eviction, and reads or writes to disk when necessary.
    In case the writethrough strategy is used, the I/O Controller directly writes to disk, 
    cache is flushed if needed and written data is added to page cache.

@@

    Our simulation model supports chunk-by-chunk file accesses
    with a user-defined chunk size. However, for simplicity, we assume that file pages are
    accessed in a round-robin fashion rather than fully randomly.
    Therefore, when a file is read, cached data is read only after all uncached data was read, and data from the inactive list is read
    before data from the active list
    (data reads occur from left to right in Figure~\ref{fig:read_order}).
    When a chunk of \emph{uncached} data is read, a new clean block is created
    and appended to the inactive list.
    When a chunk of \emph{cached} data is read, one or more existing data blocks in the LRU lists are accessed.
    If these blocks are clean, we merge them together, update the access time and size of the resulting block,
    and append it to the active list.
    If the blocks are dirty, we move them independently to the active list, to preserve their entry time.
    Because the chunk and block sizes may be different, there are situations
    where a block is not entirely read.
    In this case, the block is split into two smaller blocks and one of them is re-accessed.
    % From Tristan: not sure where to put this nor if it's necessary:
    % , in which chunks are read/written
    % until file is entirely read/written.
    \begin{figure}
           \centering
           \includegraphics[width=\columnwidth]{figures/read_order.pdf}
           \caption{File data read order. Data is read from left to right: uncached data
           is read first, followed by data from the inactive list, and finally data from the active list. }
           \label{fig:read_order}
    \end{figure}

@@

    The main simulated thread in the Memory Manager can flush or evict data from the
    memory cache. The data flushing simulation
    function takes the amount of data to flush as parameter. While
    this amount is not reached and dirty
    blocks remain in cache, this function traverses the sorted
    inactive list, then the sorted active list, and writes the
    least recently used dirty block to disk, having set its dirty
    flag to 0. In case the amount of data to flush requires that a
    block be partially flushed, the block is split into two blocks,
    one that is flushed and one that remains dirty. The time needed
    to flush data to disk is simulated by the storage model.

    The cache eviction simulation also runs in
    the main thread. It frees up the page cache by traversing and deleting
    least recently used clean data blocks in the inactive list.
    The amount of data to evict is passed as a parameter and data blocks are deleted
    from the inactive list until the evicted data reaches the required amount,
    or until there is no clean block left in the list.
    If the last evicted block does not have to be entirely evicted, the block is split into two blocks,
    and only one of them is evicted.
    The overhead of the cache eviction algorithm is not part of the simulated time
    since cache eviction time is negligible in real systems. % \tristan{a reference would be welcome}.

@@

        \end{algorithmic}
    \end{algorithm}
    Algorithm~\ref{alg:write} describes how chunk writes in
    the I/O Controller are simulated.
    % Data is written to cache until the amount of dirty data in the cache
    % exceeds the dirty ratio times the amount of available memory. Data is then written to disk.
    The algorithm initially checks the  amount of dirty data that
    can be written given the dirty ratio (line 5).
    If this amount is greater than 0, the Memory Manager is requested to evict
    data from cache if necessary (line 7).
    After eviction, the amount of data that can be written to
    page cache is calculated (line 8), and a cache write is simulated (line 9).
    If the dirty threshold is reached and there is still data to write,
    the remaining data is written to cache in a loop
    where we repeatedly flush and evict data from the cache (line 12-18).

    The above model describes page cache in writeback
    mode. Our model also includes a write function in writethrough mode,
    which simply simulates a disk write with the amount of data passed in,
    performs cache eviction if needed and adds the written data to the cache.

@@

            We first created a standalone prototype
            simulator to evaluate the accuracy and correctness of our
            model in a simple scenario before integrating it into the
            \wrench framework.
            The prototype uses the following basic storage model for
            both memory and disk:
            \begin{align*}
                & t_{r} = D / b_r \\
                & t_{w} = D / b_w\
            \end{align*}

@@

        \subsection{Experimental Methodology}

        \todo[inline]{HC: I've renamed this subsection from "Experiments" to "Experimental Methodology", as otherwise I was expecting results in there.  Not a big deal.}
        \todo[inline]{HC: I find it strange to have the experimental methodology section be a subsection of the "Methods" section, and then have the results in a new section. So I would actually make the "Experimental Methodology" a section, instead of a subsection.  The flow would be: introduction; background and related work; methods; experimental methodology; results; conclusion. Again not a big deal, just a thought.}

        Our experiments compare real executions with our Python prototype,
        with the original \wrench simulator, and with our \wrench-cache
        extension. Executions included single-threaded and multi-threaded
        applications, accessing data on local and network file systems. We
        used two applications: a synthetic one, created to evaluate the
        simulation model, and a real one, representative of neuroimaging
        data processing.

        Experiments are run on a dedicated cluster at
        Concordia University, with one login node, 9 compute nodes, and 4
        storage nodes connected with a 25 Gbps network. Each
        compute node has 2 $\times$ 16-core Intel(R) Xeon(R) Gold 6130 CPU
        @ 2.10GHz, 250~GiB of RAM, 6 $\times$ SSDs of 450~GiB each with the XFS
        file system, 378~GiB of tmpfs, 126~GiB of devtmpfs file system,
        CentOS~8.1 and NFS version 4. We use the \texttt{atop}
        and \texttt{collectl} tools to monitor and collect memory status
        and disk throughput. We clear the page
        cache before each application run to ensure comparable
        conditions.

        \begin{table}[b]
            \centering
            \begin{tabularx}{0.8\columnwidth}{c>{\centering\arraybackslash}X}
            \toprule
                Input size (GB)  & CPU time (s)\\
            \midrule
                3      & 4.4 \\
                20  & 28 \\
                50  & 75 \\
                75  & 110 \\
                100  & 155 \\
            \bottomrule
            \end{tabularx}
            \caption{Synthetic application parameters}
            \label{table:cputime}
            \end{table}
        The synthetic application, implemented in C, consists of three single-core,
        sequential tasks where each task reads the file produced by the
        previous task, increments every byte of this file to emulate real
        processing, and writes the resulting data to disk. Files are
        numbered by ascending access times (File 1 is the file read by Task 1, etc).
         The anonymous memory used by the application
        is released after each task, which we also simulate in the Python
        prototype and in \wrench-cache. As our focus is on I/O rather than compute, we measure
        application task CPU times on a cluster node
        (Table~\ref{table:cputime}), and use these durations in our
        simulations. For the Python prototype, we inject CPU times
        directly into the simulation. For \wrench and \wrench-cache, we
        determined the corresponding number of flops on a 1~Gflops CPU
        and used these values in the simulation. The simulated
        platform and application are available at
        commit \href{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}{ec6b43561b}.

        We used the synthetic application in three experiments. In the
        first one (\textit{Exp~1}), we run \emph{a single instance} of
        the application on a single cluster node, with different input file
        sizes (20~GB, 50~GB, 75~GB, 100~GB), and with all I/Os directed to
        the same local disk.
        In the second experiment (\textit{Exp~2}), we run
        \emph{concurrent} instances of the application on a single node,
        all application instances operating on different files of size 3~GB
        stored in the same local disk. We vary the number of concurrent
        application instances from 1 to 32 since cluster nodes have 32 CPU
        cores.
        In the third experiment (\textit{Exp~3}), we use the same
        configuration as the previous one, albeit reading and writing
        on a 50-GiB NFS-mounted partition of a 450-GiB remote disk of
        another compute node. As is commonly configured in HPC
        environments to avoid data loss, there is no client write cache
        and the server cache is configured as writethrough instead of
        writeback. NFS client and server read caches are enabled. 
        Therefore, all the writes happen at disk bandwidth, but
        reads can benefit from cache hits.

        The real application is a workflow of the Nighres
        toolbox~\cite{huntenburg2018nighres}, which implements cortical
        reconstruction from brain images in four steps: skull stripping,
        tissue classification, region extraction, and cortical
        reconstruction. Each step reads files produced by the previous step,
        and writes files that are or are not read by the subsequent step.
        More information on this application is available in the Nighres
        documentation at
        \url{https://nighres.readthedocs.io/en/latest/auto_examples/example_02_cortical_depth_estimation.html}.
        The application is implemented as a Python script that calls Java
        image-processing routines. We use Python 3.6, Java 8, and Nighres
        1.3.0. We patched the application to remove lazy data loading and
        data compression, which made CPU time difficult to separate from
        I/O time, and to capture task CPU times to inject in the
        simulation. The patched code is available at
        \url{https://github.com/dohoangdzung/nighres}.
        \begin{table}[t]
            \centering
            \begin{tabular}{lccc}
            \toprule
                \multicolumn{1}{c}{Workflow step}& Input size       & Output size      & CPU time\\
                                       & (MB)             & (MB)             & (s)\\
            \midrule
               Skull stripping         &  295             & 393               & 137 \\
               Tissue classification   &  197              & 1376              & 614 \\
               Region extraction       &  1376             & 885              & 76 \\
               Cortical reconstruction &  393              & 786              & 272\\
            \bottomrule
            \end{tabular} 
            \caption{Nighres application parameters}
            \label{table:nighres_stats}
            \end{table}
            \begin{table}[b]
                \centering
                \begin{tabularx}{\columnwidth}{ll
                >{\centering\arraybackslash}X
                >{\centering\arraybackslash}X
                >{\centering\arraybackslash}X}
                \toprule
                    \multicolumn{2}{c}{Bandwidths}  & Cluster (real) & Python prototype & \wrench simulator\\
                \midrule
                \multirow{2}{*}{Memory}      & read  & 6860 & 4812 & 4812\\
                                             & write & 2764 & 4812 & 4812\\
                \multirow{2}{*}{Local disk}  & read  & 510  & 465  & 465\\
                                             & write & 420  & 465  & 465\\
                \multirow{2}{*}{Remote disk} & read  & 515  & -    & 445\\
                                             & write & 375  & -    & 445\\
                \multicolumn{2}{l}{Network}  & 3000  & -    & 3000\\
                \bottomrule
                \end{tabularx}
                \caption{Bandwidth benchmarks (MBps) and simulator configurations.
                The bandwidths used in the simulations were the average of the measured read and write bandwidths.
                Network accesses were not simulated in the Python prototype.}
                \label{table:benchmark}
                \end{table}
